Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-11 paper/paper.tex
Added prose for store widening, generic fixpoint and the baseline.

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

@@

\begin{alltt}
;; Expr \(\to\) Set[State]
(define (eval e)
  (fix step (set (inj e))))
\end{alltt}

We compute the semantics of a program by iterating the state
transition relation until a fixed point in the reachable states is
reached.

\begin{alltt}
;; \(\forall\) (X) ((X \(\to\) Set[X]) \(\to\) (Set[X] \(\to\) Set[X]))
(define ((appl f) s)
  (for/fold ([i (set)])
    ([x (in-set s)])
    (\(\cup\) i (f x))))

;; Calculate fixpoint of (appl f).
;; \(\forall\) (X) ((X \(\to\) Set[X]) Set[X] \(\to\) Set[X])
(define (fix f s)
  (let loop ((accum (set)) (front s))
    (if (set-empty? front)
        accum
        (let ((new-front ((appl f) front)))
          (loop (\(\cup\) accum front)
                (new-front . \(\setminus\) . accum))))))
\end{alltt}

@@

\begin{alltt}
;; \(\widehat{State}\) = (cons (Set Conf) Store)

;; (State \(\to\) Set[State]) \(\to\) \(\widehat{State}\) \(\to\) \(\widehat{State}\)
(define ((wide-step step) state)
  (match state
    [(cons cs \(\msto\))
     (define ss ((appl step)
                 (for/set ([c cs]) (c->s c \(\msto\)))))
     (set (cons (for/set ([s ss]) (s->c s))
                (\(\sqcup\)-stores ss)))]))
\end{alltt}

@@

>>>>

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

@@

We compute the semantics of a program by iterating the state
transition relation until a fixed point in the reachable states is
reached. We begin with a generic fixpoint combinator, ${\mathcal F}$:

\begin{align*}
{\mathcal F}(f)(X) &= {\mathcal G}(f, X, \varnothing) \\
{\mathcal G}(f, \varnothing, S) &= S \\
{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup f(X)) \\
\end{align*}

The $\machstep$ relation is most naturally written as a function from
a state to a set of all next states. We want to find a fixpoint of
$\machstep$ on the state space, however, so we must lift it to work on
sets of states. The lifting is a natural one:

\begin{align*}
lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
\end{align*}

Then $eval(e) = {\mathcal F}(lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.

If we run this on our Church numeral example, we must be
patient. Indeed, the state space is finite, but it is exponential in
the size of the program. With $k = 0$, the store can be in one of ${\mathcal
  O}(2^n)$ many configurations! We take a step toward a polynomial
complexity baseline by lifting the store out so that there is one
shared amongst all states.

@@

A common technique to faster convergence in flow analyses is to share
a common store that is monotonically updated as interpretation
continues. In the ISWIM language with only unary operators, this makes
our 0-CFA quartic in the size of the program. Not great, but
polynomial.

We can reuse the reduction relation we have defined, but with some
scaffolding around it. A state in the store-widened machine is a pair
of a set of ISWIM states without a store, and a store. A step of this
machine will pair the store with each of the states in this set,
reduce, and combine the output stores to create the next. Let $wn :
\widehat{State} \times Store \to State$ be the metafunction that adds
a store to a storeless state. Formally,

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c', \msto^c), c \in cs\} \\
             \msto' &= \bigcup\{\msto^c \mid wn(c, \msto) \machstep wn(c', \msto^c), c \in cs\}
\end{align*}

@@

The final approximation we make to get to our baseline analysis is
store-allocating results of application sub-expressions. Notice that
the $\kfn{}$ continuation stores a value - this makes the space of
continuations quadratic rather than linear in 0-CFA. We instead
allocate an address for the value position when we create the
continuation. This address and the tail address are both determined by
the label of the application point, so the space becomes linear and
the overall complexity drops to cubic. Note that this is a critical
abstraction in languages with n-ary functions, since otherwise the
continuation space becomes ${\mathcal O}(n^n)$!

\begin{align*}
\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\msto,\mkont} &\machstep
\ev{\mexpi{0},\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
\\
\co{\kar{\mexp,\mlab},\mval,\msto} & \machstep
\ev{\mexp,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
\\
\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
\\
\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
\\
\ap[_\mlab]{\clos{\mvar,\mexp},\mlab^a, \msto,\mkont} & \machstep
\ev{\mexp,\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
\\
\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
\end{align*}

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
either ANF or CPS.
